Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors

### 2021-07-27 3_r_scripts/Manuscript.Rmd
Made updates and switched to SSH.

---
title: "Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors"
author: "Conor Taff"
date: "Last Updated: `r format(Sys.time(), '%B %d %Y at %H:%M')`"
output: pdf_document
bibliography: references.bib
csl: animal-behaviour.csl
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given it's role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. Although sophisticated approaches are available to partition this variation statistically, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times during a single acute response or across multiple instances of acute responses. Empiricists have largely adopted the strategy of standardizing sampling as much as possible to allow for comparison between individuals, but this standardization also makes it very difficult to detect certain types of variation in the functional shape of acute response curves. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I describe a set of simulation functions that can generate realistic acute glucorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters (e.g., the degree of correlation between the speed and scope of a response). Using this simulation, I explore several possible scenarios to highlight areas where simulation might either provide new insight into physiological flexibility directly or aid in designing empirical studies that are better able to test the hypotheses of interest. Complete code and a worked vignette using the functions are provided so that the simulation can be adapted to any study system.

@@

More recently, a series of conceptual papers have asked whether the degree of within-individual variation in glucocorticoid modulation (i.e., endocrine flexibility) across different contexts or in response to different stressors might also be an important predictor of performance [@taff2016; @hau2016; @wada2014; @lema2013]. Perhaps the major limit to empirical progress, especially for within-individual variation, is the logistical difficulty of accurately characterizing the functional shape of the acute physiological stress response for an individual during a single acute response and across multiple acute responses occurring under different conditions. Often, these measures are strictly limited by the number of samples that can safely be taken from an animal during a single capture and the number of repeated captures that are possible [but see @koolhaas2011]. Given these limitations, data simulation is a powerful tool that could complement empirical work in this area, but that has not yet been applied to studies of endocrine flexibility.

Several recent papers have suggested that physiologists interested in endocrine flexibility should adopt a within-individual reaction norm approach [e.g., @hau2016; @taff2016]. This approach has been widely adopted in studies of behavioral flexibility where statistical approaches and empirical progress have developed synergistically [e.g., @araya2015; @dingemanse2010; @westneat2015]. This field has also benefited from simulation studies to evaluate optimal study design [@van2012] and packages that can create artificial datasets with desired patterns of between, within, and residual variance to evaluate the consequences of different patterns of variation on the ability to detect effects [see SQuID package, @allegue2017]. While these approaches are powerful, they have proven difficult to apply directly to endocrine flexibility data for two reasons. First, simulation studies suggest that many patterns may only be detectable with a level of repeated sampling that is possible for many behaviors (especially when collected autonomously), but that is currently not possible for most studies of endocrine flexibility. Second, and more fundamentally, these papers often focus on somewhat discrete measures of behavior (e.g., aggression score or activity level), whereas for acute glucocorticoid responses, the functional shape of the response itself may be the important trait and it may not be possible to summarize variation in the shape of the response with a single measure.

@@

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017].

@@

Data simulation is a powerful approach for several reasons. Because true parameter values (e.g., maximum glucocorticoid level) are known, it is possible to evaluate how well different study designs and analytical choices perform in recovering true patterns and how sensitive those designs are to different assumptions. Thus, simulation can tell us whether the study designs we use can *in principle* detect the patterns we predict given realistic effect sizes. Simulated data can also identify conditions under which current study designs will perform well or poorly. For example, if simulations suggest that the baseline paired with stress-induced paradigm only works well when the speed and scope of responses are positively correlated, then empirical work could seek to determine the degree of correlation for a particular study system as justification for the approach. This ability to highlight key assumptions and create data sets with known properties has the ability to both provide insight into physiological flexibility directly and to guide empirical work by improving study design and identifying key areas for subsequent sampling. In the rest of this paper, I develop a simple simulation of acute physiological stress responses and then briefly illustrate several possible applications of the simulation.

@@

I developed a set of simulation functions in R version 4.0.2 [@rcore] to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure 1). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response.

![](concept.png)\

***Figure 1.** Conceptual illustration of the structure of the simulation. For each simulated animal, seven parameters are sampled from a multivariate normal distribution. Together, these seven parameters define the turning points in an acute response curve. The mean and standard deviation for each parameter can be set along with the degree of covariation between each pair of parameters. Note that the simulation can easily be simplified as desired by setting some parameter mean or standard deviations to zero.*

The simulation is constructed as two main functions with several minor functions for downstream analysis. I have included a detailed description of the functions and arguments in the appendix along with a full vignette detailing the process of specifying values for each function (https://github.com/cct663/speed_vs_scope). Briefly, function `cort_sim1` samples the parameters shown in Figure 1 from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. I consider these values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). Second, additional noise can be added to all measurements to represent measurement error (e.g., assay error, inaccurate pipetting, etc). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response.

The function also generates a simulated performance (e.g., fitness) measure, based on the underlying true values. In addition to returning the full curve, a down sampled dataset is returned that retains measurements made only at the user specified fixed time points (e.g., only at 2 and 30 minutes, as is typical in many field based studies). Data generated from this function can then be used in downstream analyses with any standard statistical approaches or software. For example, a user could perform an analysis to ask whether a known relationship between fitness and a particular true parameter is recovered in an analysis that includes only measures taken at particular time points. An additional convenience function summarizes the output of a simulation run in a multi-panel plot (Figure 2).

![](demo.png)\

***Figure 2.** Example of simulation output with default settings. Panel A shows the down sampled data set for this run with samples collected at 1, 15, and 30 minutes in this case. Panel B shows the full observed response curve for each animal. Panel C shows the rank order of glucocorticoid level at each time point for each animal. In each panel, the vertical dashed lines represent the three time points that might have been measured in a typical empirical study.*

Finally, given recent interest in estimating the repeatability of glucocorticoid regulation [@taff2018; @cockrem2013; @hau2016], I also included a function that takes input from `cort_sim2` and calculates the observed repeatability of several measures using package `rptR` [@stoffel2017]. Full details are included in the appendix, but this function returns repeatability for each individual time point specified in the down sampled data set, profile repeatability [@reed2019], and for area under the curve calculated as both increase (AUC~I~) and ground (AUC~G~) approaches [@pruessner2003]. For each AUC measure, I return repeatability for the full time course, for an estimate using only the observed values in the down sampled data set, and for the full data set constrained to the time period encompassing the observed data points.

@@

@@

### *Comparing repeatability of different measures*

>>>>

---
title: "Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
fontsize: 12pt
output: 
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(ggplot2, here, MASS, MBESS, tidyverse, viridis, gridExtra, rptR, simcoRt, sp)
```

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given it's role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. Although sophisticated approaches are available to partition this variation statistically, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times during a single acute response or across multiple instances of acute responses. Empiricists have largely adopted the strategy of standardizing sampling as much as possible to allow for comparison between individuals, but this standardization also makes it very difficult to detect certain types of variation in the functional shape of acute response curves. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I describe the `simcoRt` package, which includes functions that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters (e.g., the degree of correlation between the speed and scope of a response). Using this simulation, I explore several possible scenarios to highlight areas where simulation might either provide new insight into physiological flexibility directly or aid in designing empirical studies that are better able to test the hypotheses of interest. The package includes a vignette with details on use and code to reproduce all of the examples explored in this paper.

@@

More recently, a series of conceptual papers have asked whether the degree of within-individual variation in glucocorticoid modulation (i.e., endocrine flexibility) across different contexts or in response to different stressors might also be an important predictor of performance [@taff2016; @hau2016; @wada2014; @lema2013]. Perhaps the major limit to empirical progress, especially for within-individual variation, is the logistical difficulty of accurately characterizing the functional shape of the acute physiological stress response for an individual during a single acute response and across multiple acute responses occurring under different conditions. Often these measures are strictly limited by the number of samples that can safely be taken from an animal during a single capture and the number of repeated captures that are possible [but see @koolhaas2011]. Given these limitations, data simulation is a powerful tool that could complement empirical work in this area, but that has not yet been applied to studies of endocrine flexibility.

Several recent papers have suggested that physiologists interested in endocrine flexibility should adopt a within-individual reaction norm approach [e.g., @hau2016; @taff2016]. This approach has been widely adopted in studies of behavioral flexibility where statistical methods and empirical progress have developed synergistically [e.g., @araya2015; @dingemanse2010; @westneat2015]. This field has also benefited from simulation studies to evaluate optimal study design [@van2012] and packages that can create artificial datasets with desired patterns of between, within, and residual variance to evaluate the consequences of different patterns of variation on the ability to detect effects [see SQuID package, @allegue2017]. While these approaches are powerful, they have proven difficult to apply directly to endocrine flexibility data for two reasons. First, simulation studies suggest that many patterns may only be detectable with a level of repeated sampling that is possible for many behaviors (especially when collected autonomously), but that is currently not possible for most studies of endocrine flexibility. Second, and more fundamentally, these papers often focus on somewhat discrete measures of behavior (e.g., aggression score or activity level), whereas for acute glucocorticoid responses, the functional shape of the response itself may be the important trait and it may not be possible to summarize variation in the shape of the response with a single measure.

@@

For individual based studies, the most common approach to this problem is to standardize measurements as much as possible by measuring animals at the same time of the day during the same context, and by taking blood samples at standard times (often <3 and 30 minutes after capture) to characterize baseline and stress-induced glucocorticoids. This standardization allows for comparison between individuals, but in some cases it may also completely obscure the ability to detect variation in certain characteristics of the acute response curve. For example, if the speed (rate of initial increase) and scope (maximum value) of the acute response vary independently, samples taken at only two time points cannot accurately capture variation in either parameter. Indeed, several discussions in recent years about methods such as the '3 minute rule' and the relative merits of 'area under the curve' versus time point measures of glucocorticoids are fundamentally related to a recognition of the importance of understanding variation in the functional shape of stress responses and whether different components of that shape covary within individuals [e.g., @cockrem2002; @small2017].

@@

Data simulation is a powerful approach for several reasons. Because true parameter values (e.g., maximum glucocorticoid level) are known, it is possible to evaluate how well different study designs and analytical choices perform in recovering true patterns and how sensitive those designs are to different assumptions. Thus, simulation can tell us whether the study designs we use can *in principle* detect the patterns we predict given realistic effect sizes. Simulated data can also identify conditions under which current study designs will perform well or poorly. For example, if simulations suggest that the baseline paired with stress-induced paradigm only works well when the speed and scope of responses are positively correlated, then empirical work could seek to determine the degree of correlation for a particular study system as justification for the approach. This ability to highlight key assumptions and create data sets with known properties has the potential to both provide insight into physiological flexibility directly and to guide empirical work by improving study design and identifying key areas for subsequent sampling. In the rest of this paper, I develop a simple simulation of acute physiological stress responses and then briefly illustrate several possible applications of the simulation.

@@

I developed a package called `simcoRt` in R version 4.0.2 [@rcore] with functions to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure \@ref(fig:concept-fig)). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response. The package can be installed in R using the following command.

```{r simp-ackage, message = FALSE, warning = FALSE}
devtools::install_github("cct663/simcoRt")
```

```{r concept-fig, echo = FALSE, message = FALSE, warning = FALSE, out.width = '90%', fig.cap = "Conceptual illustration of the structure of the simulation. For each simulated animal, seven parameters are sampled from a multivariate normal distribution. Together, these seven parameters define the turning points in an acute response curve. The mean and standard deviation for each parameter can be set along with the degree of covariation between each pair of parameters. Note that the simulation can easily be simplified as desired by setting some parameter mean or standard deviations to zero."}
knitr::opts_chunk$set(fig.pos = "!H")

knitr::include_graphics("concept.png")

```

The simulation is constructed as two main functions with several minor functions for downstream analysis. Detailed descriptions of the arguments to each function and a full vignette are included with the package documentation. Briefly, function `cort_sim1` samples the parameters shown in Figure \@ref(fig:concept-fig) from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. I consider these values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response. From this expressed response, individual data points are then collected at user specified times that would reflect an empirical study design (e.g., 2, 30, and 60 minutes). Additional noise is added to these data points to represent measurement error (e.g., assay error).

The function also generates a simulated performance (e.g., fitness) measure, based on the underlying true values. Data reflecting the true phenotypic values, the repeated expression of acute responses, and the observed time points can then be used in downstream analyses with any standard statistical approaches or software. For example, a user could perform an analysis to ask whether a known relationship between fitness and a particular true parameter is recovered in a study that includes only measures taken at particular time points. An additional convenience function summarizes the output of a simulation run in a multi-panel plot (Figure \@ref(fig:initial-demo)).

```{r initial-demo, results = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10.2, fig.height = 7.5, fig.cap = "Example of simulation output with default settings. Panel A shows the downsampled data set for this run with samples collected at 1, 15, and 30 minutes in this case. Panel B shows the full observed response curve for each animal. Panel C shows the rank order of glucocorticoid level at each time point for each animal. In each panel, the vertical dashed lines represent the three time points that might have been measured in a typical empirical study. Note that individuals in the top panels do not match perfectly because measurement error is added to the downsampled dataset in panel A."}
  # Make plot with default settings from package 
     set.seed(100)   # makes it reproducible by using same random seed
      demo <- cort_sim2()
      p <- plot_cort_sim(demo)
      p
```

Finally, given recent interest in estimating the repeatability of glucocorticoid regulation [@taff2018; @cockrem2013; @hau2016], I also included a function that takes input from `cort_sim2` and calculates the observed repeatability of several measures using package `rptR` [@stoffel2017]. Full details are included in the package documentation, but this function returns repeatability for each individual time point specified in the down sampled data set, profile repeatability [@reed2019], and repeatability for area under the curve calculated as both increase (AUC~I~) and ground (AUC~G~) approaches [@pruessner2003]. For each AUC measure, the function returns repeatability for the full time course, for an estimate using only the observed values in the down sampled data set, and for the full data set constrained to the time period encompassing the observed data points. Simple plots illustrating repeated samples from the same individuals are also returned by default.

@@

In order for simulation to be useful, we should be able to create artificial datasets that have similar characteristics to empirical data for different systems. Simulating realistic data provides a starting point for evaluating different study designs and the consequences of changes in different assumptions or parameters. Simulating realistic data is also useful because it can aid in study design or be used as a basis for pre-registered reports that demonstrate the feasibility of a planned study before data are ever collected. Simulated data can be created and entered in a complete analysis pipeline, with empirical data substituted later. In addition to helping to design better studies, this approach has the advantage of increasing the transparency and reliability for studies of physiological flexibility, by making analysis choices and predictions clear before data are collected.

Here, I compare simulated data to an empirical dataset from red-winged blackbirds (*Agelaius phoeniceus*) to demonstrate how `simcoRt` can create synthetic datasets similar to observed data. This dataset is included in the `simcoRt` package and is fully described in the documentation. Briefly, acute glucocorticoid responses were measured using a 7-point sampling scheme, with corticosterone measured at 1, 3, 7, 15, 30, 45, and 60 minutes post capture. I simulated data using the functions described above starting with the input values calculated directly from the empirical data. The simulation creates a new dataset that has similar variation and patterns to empirical data (Figure \@ref(fig:rwbb-demo)A) along with a population wide corticosterone response curve shape that closely matches the empirical data (Figure \@ref(fig:rwbb-demo)B). In this case, the plotted simulation data include the same number of animals sampled at the same time points as the empirical data, but these sampling points and total sample size can easily be changed as desired. The parameterized simulation can now be used to test the sensitivity of any number of experimental designs before additional data is collected.

```{r rwbb-demo, echo = FALSE, fig.height = 3.5, fig.width = 8, message = FALSE, warning = FALSE, fig.cap = "Panel A shows the acute corticosterone response for measured (blue) or simulated (orange) red-winged blackbirds measured at seven time points. Panel B shows the mean and standard error of the two datasets."}
      rwde <- subset(rwbb, rwbb$lh_stg == "early-breeding")  
      rwde <- subset(rwde, rwde$id != "452")   # This one has an unusually high cort response
    set.seed(165)
        sim_rwd <- cort_sim2(cort_sim1(
       n = 23,
       base_mu = 3.5,
       base_sd = 3.5,
       slope_mu = 17,
       slope_sd = 0.7,
       fastpct_mu = 0.7,
       fastpct_sd = 0.03,
       max_mu = log(56),
       max_sd = 0.5,
       speed_mu = 65,
       speed_sd = 5,
       return_mu = 120,
       maxtime_mu = 25,
       cor_max_slope = -0.7
     ), 
     sample_times = 1,
     bleed_times = c(1, 3, 8, 15, 30, 45, 60),
     assay_error = 0.2,
     sm_span = 0.2)

     sim_rwd$simulated_dataset_long$type <- "simulated"
     sim_rwd2 <- sim_rwd$simulated_dataset_long[, c("animal_sample", "time", "cort", "type")]
     rwde2 <- rwde[, c("id", "lat_sec", "cort")]
     rwde2$lat_sec <- rwde2$lat_sec / 60
     rwde2$type <- "red-winged blackbird"
     colnames(rwde2) <- c("animal_sample", "time", "cort", "type")

     comp_sim <- rbind(rwde2, sim_rwd2)

     p1 <- ggplot(comp_sim, mapping = aes(x = time, y = cort, by = animal_sample, color = type)) +
       geom_line(alpha = 0.7, size = 0.7) +
       theme_classic() +
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 125)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       guides(color = FALSE) +
       theme(axis.title = element_text(size = 14)) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "A")

     p2 <- ggplot(comp_sim, mapping = aes(x = time, y = cort, color = type, fill = type)) +
       geom_smooth() +
       theme_classic() + 
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 125)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       scale_fill_manual(values = c("slateblue", "orange")) +
       theme(legend.position = c(0.7, 0.15), axis.title = element_text(size = 14)) +
       guides(color = guide_legend(title = ""), fill = guide_legend(title = "")) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "B")

      ggpubr::ggarrange(p1, p2, nrow = 1) 
```

### *Accurately measuring a single glucocorticoid trait*

Single time point measures of glucocorticoids are often interpreted as representing meaningful variation between individuals. For example, variation in the level of glucocorticoids after 30 minutes of standardized restraint is typically interpreted as variation in the magnitude of the stress response [@taff2019]. However, this interpretation rests on assumptions that are rarely explicitly tested with empirical data. For example, the time chosen to take a stress-induced sample is often assumed to be either at the species peak or during a plateau period after the species peak. In some early studies, great care was taken to determine an average population level peak time [CITE], but many studies adopt the widely used 'standard' time of 30 minutes post capture without extensive validation. While there is a general assumption that sampling later than the peak is acceptable (and perhaps preferable) because animals will be sampled during a relatively stable high plateau, there is little empirical data to evaluate this assertion or to determine how much under or overshooting the species peak timing might influence inferences. Furthermore, even when the average peak timing is well established, differences in the amount of between-individual variation in the time to reach the peak or in peak values are common across species and even in different life history stages within species [CITE]. The combinations of these patterns of variation could have major consequences on the accuracy of single point estimates taken at 30 minutes, but these questions cannot be addressed directly with empirical datasets where the true underlying values of each individual are unknown.

Here, I simulate a simple scenario exploring the consequences of variation in each of these parameters on the accuracy of estimating between individual differences in maximally expressed glucocorticoids during an acute response. For purposes of this illustration, I consider a single study design in which animals are sampled at 30 minutes. Using this design as a starting point, I systematically vary i) the timing of the population average peak (15, 30, or 45 minutes), ii) the amount of variation in maximum glucocorticoid levels reached, iii) and the amount of variation in the number of minutes taken to reach peak levels. All other variables in the simulation are constrained to be invariant between individuals in the population (e.g., all individuals have identical baseline glucocorticoids in this case), though I consider cases in which multiple aspects of the rapid response are correlated with each other in the next section. I included moderate within-individual variability and a small amount of assay error across all iterations. For each combination of parameters, I simulated 200 animals and estimated the R^2^ value from a regression of the observed estimates of glucocorticoid levels at 30 minutes to the true known values. This simulation is likely a best case scenario because it eliminates many sources of variation or noise that would be present in real data, but it illustrates the effect of variation in these three key parameters even when the exact same sampling design is employed.

```{r max-sim, echo = FALSE, warning = FALSE, message = FALSE, fig.pos = "!h", fig.align = "center", fig.width = 8, fig.height = 3, fig.cap = "Results of simulation runs with different amounts of between-individual variation in the time to reach maximum glucocorticoid levels and in the maximum level reached. Simulations are run with samples taken at 30 minutes on populations with an average peak time of 15 minutes (left), 30 minutes (center), or 45 minutes (right). Each grid cell is the R$^2$ value from the regression of observed glucocorticoids at 30 minutes to true maximum levels in a simulation of 200 individuals."}
# set.seed(901)
# grid.size <- 15
# 
# max_sd <- seq(0.01, 0.26, length.out = grid.size)
# time_sd <- seq(1, 20, length.out = grid.size)
# #plateau <- c(2, 20)
# speeds <- c(15, 30, 45)
# 
# storer <- data.frame(max_sd = rep(rep(max_sd, grid.size), 3),
#                      time_sd = rep(rep(time_sd, each = grid.size), 3),
#                      rsq = rep(rep(NA, grid.size*grid.size), 3),
#                      slp = rep(rep(NA, grid.size*grid.size), 3),
#                      #plateau = rep(c("Short plateau", "Long plateau"), each = 300),
#                      speeds = rep(c("Mean max 15m", "Mean max 30m", "Mean max 45m"), each = grid.size*grid.size))
# 
# counter <- 0
# 
# #for(u in 1:length(plateau)){
#   for(k in 1:length(speeds)){
#     for(i in 1:length(time_sd)){
#         for(j in 1:length(max_sd)){
#           dm <- cort_sim2(data = cort_sim1(n = 200, base_mu = 0, base_sd = 0, slope_sd = 0, slope_mu = 5, fastpct_mu = 0.4, fastpct_sd = 0, maxtime_sd = 0, return_sd = 0,
#                                          speed_mu = speeds[k], 
#                                          speed_sd = time_sd[i],
#                                          max_mu = log(45),
#                                          max_sd = max_sd[j],
#                                          maxtime_mu = 10,
#                                          return_mu = speeds[k] + 60
#                                          ),
#                         bleed_times = c(2, 30),
#                         sample_times = 1)
#                         #base_error = 0, speed_error = 0, max_error = 0, maxtime_error = 0, return_error = 0, slope_error = 0, fastpct_error = 0)
#         
#           dm2 <- data.frame(max30 = subset(dm$simulated_dataset_long$cort, dm$simulated_dataset_long$time == 30),
#                           max_true = dm$true_values$max)
#           m <- lm(max_true ~ max30, data = dm2)
#           rs <- summary(m)$r.squared
#           
#           storer[counter + j, 3] <- rs
#           storer[counter + j, 4] <- coef(m)[2]
#         }
#       counter <- counter + grid.size
#       print(paste(i, "of", length(time_sd), sep = " "))
#     }
#     print(paste(k, "of", length(speeds)))
#   }
# #  print(paste(u, "of", length(plateau)))
# #}

#saveRDS(storer, here::here("5_other_outputs/max_sim.rds"))
storer <-readRDS(here::here("5_other_outputs/max_sim.rds"))

# Making bins for r squared values
  storer$rsq2 <- cut(storer$rsq, breaks = seq(0, 1, 0.2))
  new_bins <- data.frame(rsq2 = levels(storer$rsq2),
                         rsq3 = c("0.0 - 0.2", "0.2 - 0.4", "0.4 - 0.6", "0.6 - 0.8", "0.8 - 1.0"))
  storer <- plyr::join(storer, new_bins, "rsq2")

# Putting max sd onto interpretable scale
  storer$max_sd2 <- (exp(storer$max_sd + rep(log(45), nrow(storer))) - exp(rep(log(45), nrow(storer)) - storer$max_sd)) / 2

ggplot(data = storer, mapping = aes(x = time_sd, y = max_sd2, fill = rsq3)) +
  geom_tile() +
  scale_fill_viridis(option = "plasma", discrete = TRUE) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
  ylab("Variation in max (SD ng/\U00B5l)") +
  xlab("Variation in time to reach max (SD minutes)") +
  facet_wrap(~ speeds, nrow = 1) +
  #facet_grid(rows = vars(plateau), cols = vars(speeds)) +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 12)) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(fill = expression(paste("R"^{2})))

```

Results of this simulation are summarized in figure \@ref(fig:max-sim). The amount of between-individual variation in the maximum glucocorticoid value has a profound effect on the ability to detect true maximal levels with samples taken at 30 minutes. In one sense, this result is unsurprising because it is intuitive that large differences would be easier to detect, but there are important consequences of this fact for interpreting studies that seek to link between-individual variation in the magnitude of the stress response with other traits. For example, the magnitude of the acute stress response often varies substantially across life history stages [CITE]. Even if study designs are identical and maximum glucocorticoids are associated with performance, it will be easier to detect those patterns during life history stages with greater variation (see section on detecting fitness associations below). There is a weaker, but still substantial impact of variation in the time taken to reach maximum values on the accuracy of estimates in this simulation. Greater variation in time reduces the accuracy of estimates of maximal values. Finally, the timing of sampling relative to the average population peak timing also influences accuracy. Measuring after the average peak time results in the most accurate estimates across a range of parameter values, while measuring before the average peak time produces the least accurate measures, particularly when there is also high variation in the time to reach maximum values between individuals. This simple example demonstrates clearly that the same experimental design will perform better or worse depending on the combination of glucocorticoid regulation parameters in the population being studied.

@@

In reality, fully characterizing the acute glucocorticoid response requires more than identifying just the maximum value reached. Individuals may differ in baseline levels, rate of initial increase, the speed of reaching the maximum level, time spent at maximum, and the speed of return to baseline. Moreover, each of these components of the endocrine response could be positively or negatively correlated with each other within and between individuals. In these cases, measurements taken at particular time points contain information about multiple aspects of the response and without additional information it may be difficult to know what trait is being measured. The fact that each of these traits might be important and that they might covary has been discussed in a general sense [CITE], but simulations are uniquely powerful for exploring under exactly what conditions time point measure of glucocorticoids can or cannot be used as indicators of these traits.
